\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-05-25 fluB.tex
More small notation fixes.

where $\chi^{2}$ is calculated from a classical contingency table, $N$ is the number of haplotypes and $(k-1)(m-1)$ are the degrees of freedom.
This statistic is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \citep{zhao2005}.
LD was estimated only at loci where each nucleotide or amino acid allele was present in at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
In all cases we used a minor allele frequency cutoff of 1\%.
We also calculated another LD statistic, $D'$ \citep{lewontin1964}:

\begin{equation}
D'_{ij}=\frac{D_{ij}}{D_{ij}^{max}},
\end{equation}

where

\begin{equation}
D_{ij}=p(A_i B_j) - p(A_i) p(B_j)
\end{equation}

and

\begin{equation}
\begin{split}
D_{ij}^{max}=min[ p(A_i) p(B_j) , (1-p(A_i))(1-p(B_j)) ] \mbox{ when } D_{ij} < 0\\
D_{ij}^{max}=min[ (1-p(A_i)) p(B_j) , p(A_i)(1-p(B_j)) ] \mbox{ when } D_{ij} \geq 0,
\end{split}
\end{equation}

where $p(A_i)$ is the frequency of allele $A_i$ at locus A, $p(B_j)$ is the frequency of allele $B_j$ at locus B and $p(A_i B_j)$ is the frequency of haplotype $A_i B_j$.
$D'$ is inflated when some haplotypes are not observed \textit{e.g.} when the minor allele frequency is low.
We find that $D'$ is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.

>>>>

where $\chi^{2}$ is calculated from a classical contingency table, $N$ is the number of haplotypes and $(k-1)(m-1)$ are the degrees of freedom.
This statistic is equal to the widely used $r^2$ LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus \citep{zhao2005}.
LD was estimated only at loci where each nucleotide or amino acid allele was present in at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
In all cases we used a minor allele frequency cutoff of 1\%.
We also calculated another LD statistic, $D'$ \citep{lewontin1964} as $D'_{ij}=D_{ij}/D_{ij}^{max}$,
where $D_{ij}=p(A_i B_j) - p(A_i) p(B_j)$ and
\begin{equation}
\begin{split}
D_{ij}^{max}=min[ p(A_i) p(B_j) , (1-p(A_i))(1-p(B_j)) ] \mbox{ when } D_{ij} < 0\\
D_{ij}^{max}=min[ (1-p(A_i)) p(B_j) , p(A_i)(1-p(B_j)) ] \mbox{ when } D_{ij} \geq 0,
\end{split}
\end{equation}
where $p(A_i)$ is the frequency of allele $A_i$ at locus A, $p(B_j)$ is the frequency of allele $B_j$ at locus B and $p(A_i B_j)$ is the frequency of haplotype $A_i B_j$.
$D'$ is inflated when some haplotypes are not observed \textit{e.g.} when the minor allele frequency is low.
We find that $D'$ is almost uniformly high across the influenza B virus genome and close to 1.0 for almost any pair of polymorphic loci.
This is because most amino acid alleles in the dataset exist transiently, meaning that they don't get a chance to reassort and we only observe them within the backgrounds of more persistent alleles, which $D'$ quantifies as complete LD.
We think that metrics related to $r^2$, like $\chiSq$, perform much better on temporal data such as ours in finding persistent associations between alleles and are easier to interpret.
